M.Tech Thesis

### 2023-07-03 build/latex/mtechthesis.tex
publishing

\usepackage{sphinxmessages}
\setcounter{tocdepth}{2}

@@

\sphinxAtStartPar
A side effect of the rapid progress in realtime audio computation is that may techniques
used principally in the context of \sphinxstyleemphasis{non\sphinxhyphen{}realtime} computer music have largely fallen out of the public light.
However, in the last decade or so computers have reached speeds such that some of the
tools and approaches previously associated only with offline rendering have become
feasible even in a realtime scenario.
One of these, the one with which I am involved, is the programming of computer music in the
Lisp family of languages.
Lisp has a rich history in computer music, going back to the very early days of the field.
With its support for symbolic processing, recursion, and iteractive development,
Lisp has always been an elegant and productive way to represent and build music.
However, the accepted wisdom has been that one does not try to do anything
realtime heavy, where heavy means the computer is generating significant amounts of digital audio,
while running a Lisp or similar high\sphinxhyphen{}level, garbage\sphinxhyphen{}collected language.
This is now not just possible, but practical even at latencies usable by serious instrumentalists
(i.e., sub 20 ms), allowing it be used on stage as well as in the studio.

\sphinxAtStartPar
As a result, I believe we are seeing a renaissance of the use of dynamic programming languages in music.
JavaScript is used as an extension language in Max.
Python, Ruby, and Clojure are all used for algorithmic composition.
Even Haskell is used in live\sphinxhyphen{}coding performances. And a new generation of computer
musicians is discovering the elegance, the productivity, and the joy of doing this work
with Lisp, a language that I and many others will argue is uniquely suited to expressing musical
abstractions in code.

\sphinxAtStartPar
The Scheme for Max project is, among other things, an attempt to make music programming in a Lisp
\sphinxhyphen{} specifically Scheme Lisp \sphinxhyphen{} accessible and practical to the “regular” computer musician working in a “regular” studio.
Scheme for Max (S4M) provides an extension to the Max audio/visual programming platform
that embeds a Scheme interpreter, allowing the user to create fully function Scheme programs that
interact with the rest of the Max environment, and which do so with high temporal accuracy.
S4M can be used to embed single\sphinxhyphen{}line programs directly in visual Max patches, or to build
programs of thousands of lines that do everything from sequencing MIDI
to controlling analog hardware and editing audio.
And it can be used for everything from complex contemporary scored music, to interactive
live improvisation, to generative and stochastic algorithmic composition.

\sphinxAtStartPar
While I am far from the only one working on new ways to use Lisp in computer music (even in a realtime context),
I believe Scheme for Max achieves something unique.
By creating an enviroment thats runs seamlessly and flexibly inside the tremendously successful
Max platform (and thus also inside the even more tremendously successful Ableton Live digital audio workstation
which is able to host Max),
Scheme can comfortably be used in close tandem with both popular,
commercial music tools and the vast library of specialized, academic, and research tools available for Max.
And by the happy coincidence of Scheme’s syntax compability with Max message syntax, it can be even be used
in baby\sphinxhyphen{}steps \sphinxhyphen{} new programmers can add the power of Scheme to their Max patches with
only a few lines of code, and that without even reaching for a text editor.
I believe Scheme’s consistent, minimal syntax and interactive development style provide an ideal
introduction to textual programming for the curious Max user.
With Scheme for Max, one is not forced to choose between the power and flexibility of Lisp and the
convenience and accessibility of commercial music software.

\sphinxAtStartPar
This thesis introduces the Scheme for Max project. I begin by discussing the landscape into which it fits
and its precendents and inspirations. I cover the motivations for the project and the goals by which
I judge its success. I then examine its design and capabilities, and finally, I evaluate
how well I feel it has succeeded and discuss my future plans for it.

\sphinxAtStartPar
For reasons of space, I have not covered every feature of the project, nor have I delved into the
low\sphinxhyphen{}level implementation details and C code. I have limited myself to discussion appropriate
to an interested computer musician, rather than a professional programmer or computer scientist.
That said, the project is open\sphinxhyphen{}source and freely available, so the
the interested reader is encouraged to consult the GitHub project page for further details if desired,
where they will find links to comprehensive documentation, help files, demo videos, and source code.
Code samples presented here in Scheme should be comprehensible to a new programmer,
but if the reader requires more help, I would point them to my “Learn Scheme For Max” online
e\sphinxhyphen{}book in which I introduce Scheme programming from first\sphinxhyphen{}principles, available again
from the main project page at (\sphinxurl{https://github.com/iainctduncan/scheme-for-max}).

@@

\sphinxAtStartPar
To begin, I would like to clarify exactly what I mean by “computer music programming”.
I am using the term to refer to programming in which both the tools used and the composition itself
are programmed on a computer.
I say this using the word “composition” in its broadest sense \sphinxhyphen{} the composition could
be a linearly scored piece of set length and content, or an interactive program with which
a performer interacts, but that contains prepared material of some kind.
This is distinct, for example, from commercial music tools (such as sequencers or digital
audio workstations) in which a computer program is run as the host environment, but the compositional
work itself is stored as simply data within that program (for example in a binary format or
in MIDI files), and would not normally be considered a computer \sphinxstyleemphasis{program} itself.

\sphinxAtStartPar
Using this definition of computer music programming, we can draw a further distinction
between two kinds of programming and programmer:
we have programming that is specific to a composer, or one or more artistic pieces of work;
and we have programming of tools and frameworks that will be potentially be used by many
people and in potentially many contexts.
For the purpose of this discussion, I will refer to the people doing these as the “composer\sphinxhyphen{}programmer”
\sphinxhyphen{} a term also used by Curtis Roads in “Composing Electronic Music” (Roads, 2015, 341) \sphinxhyphen{}
and the “tools programmer” (my own terminology).
Of course, these frequently are the same person at different times.
This distinction is important, as the goals of these two programmers, and thus the
ideal design of supporting technology, can frequently be very different.
The composer\sphinxhyphen{}programmer likely wishes the programming tools to be as immediately convenient as possible to
an individual, while the tools\sphinxhyphen{}programmer may be
developing software for use by a larger community and may prefer design decisions
that favour long\sphinxhyphen{}term code reuse and team development at the expense of immediate convenience.
As we will see, the various computer music languages, platforms, and approaches
favour one of these hypothetical programmers to a greater or lesser degree.

\sphinxAtStartPar
Beyond these distinctions, I use the term “computer music programming” as broadly
as possible \sphinxhyphen{} I include as computer music programming platforms both graphic platforms with
which a beginner can be immediately productive as well as tools for advanced programming
languages that are only accessible to experienced computer programmers.

\section{Requirements for a Computer Music Platform}
\label{\detokenize{background:requirements-for-a-computer-music-platform}}
\sphinxAtStartPar
Prior to examining the programming platforms, let us examine some
common requirements of the composer\sphinxhyphen{}programmer so that we can assess how the various
options meets these needs.
While the list of possible requirements is potentially long, I will group
them into three high level requirements that are personally important to me as
a composer and performer. Note that not all of these are necessarily fulfilled well, or at all,
by each of the possible platforms a composer might use \sphinxhyphen{} some offer
only partial support for a requirement, or perhaps even none at all (Lazzarini, 2013, 1).

\subsection{Support for Musical Abstractions}
\label{\detokenize{background:support-for-musical-abstractions}}
\sphinxAtStartPar
To be maximally productive in producing musical works, the computer music language should provide
musically meaningful abstractions that support how the composer thinks about music.
Examples of these abstractions would be programmatic constructs for representing
notes, scores, instruments, voices, tempi, bars, beats, and the like.
The more of these that are supported in the language itself, the less the programmer
must decide and code themselves in order to make music.
Probably the most important of these abstractions are those concerning musical time
\sphinxhyphen{} a good computer music language must have an implementation of \sphinxstyleemphasis{logical time},
allowing the programmer to schedule events accurately in a way that makes sense
to the user but that also works for rendering or playing in real\sphinxhyphen{}time to a degree of accuracy
appropriate for music (Dannenberg, 2018, 3).

\sphinxAtStartPar
However, it should be noted that the importance of these abstractions being readily provided
by a language varies widely amongst composer\sphinxhyphen{}programmers:
while some composers will be relieved not to have to decide how to implement
the concept of beats and tempi themselves, others may be expressly looking for platforms
that do not come with any assumptions of how one might think about musical abstractions.
Computer music programming languages thus also vary widely in their areas
of focus and the problems they attempt to solve for the programmer (Dannenberg, 2018, 4).
.. citation (Dannenburg, 2018)
For example, the Csound language comes with a built\sphinxhyphen{}in abstraction of a score system,
that in turn depends on abstractions representing beats and tempi (Lazzarini, 2016, 157).
.. citation (Lazzarini, 2016)
In contrast, the Max platform takes the design approach of giving the user
an almost blank canvas, deliberately avoiding any assumptions that the user
will make music of any particular style or in any particular way (Puckette, 2002, 34).
.. citation (Puckette, 2002)

@@

\sphinxAtStartPar
Again, the importance of this varies with the composer and the platform.
In fact, this was not even possible with early computer music languages such
as the MUSIC\sphinxhyphen{}N family, with which the process of rendering audio from a score file
took far longer than the duration of a piece (Wang, 2008, 6).
.. citation (Wang, 2008) needs page checked  FIX

\sphinxAtStartPar
However, since the advent of realtime\sphinxhyphen{}capable systems, this has become a
standard feature in computer music platforms.
In fact, in contrast to the Music\sphinxhyphen{}N languages, the Max platform was designed
to support performance\sphinxhyphen{}time interactivity first, with the ability to render audio only added later
as computers became capable of doing both (Puckette, 2002, 33).
.. citation (Puckette, 2002)
As someone who comes principally from a jazz and contemporary improvised music
background, being able to create complex systems that support realtime input is a high priority to me.

\subsection{Support for Complex Algorithms}
\label{\detokenize{background:support-for-complex-algorithms}}
\sphinxAtStartPar
Finally, the platform should provide support for programming algorithms of significant complexity.
It should be possible to make the platform, for example, do things that are impractical
or impossible for the human perfomer or for the composer working at a piano with score paper.
My personal goals include creating music that combines algorithmically generated
material with traditionally scored and improvised material.
For my algorithmic work, I want to explore techniques and forms that would be impractical
if one is individually entering notes, thus instead requiring programmatic implementation of
algorithms.
However, it is also important to me that a performer be able to interact with the algorithms.
That, to quote Miller Puckette on his original design goals for Max,
basic musical material should be “controllable physcially, sequentially, or algorithmically;
if they are to be controlled algorithmically, the inputs to the algorithm should
themselves be controllable in any way” (Puckette, 1991, 68).
.. citation (Puckette, 1991)

@@

\sphinxAtStartPar
I will briefly discuss each of these, listing examples, but focusing on a representative tool from each family.
I will provide my observations and experiences of the advantages and disadvantages of each,
drawing both on the literature and on my personal experiences with tools from each category
over the last 25 years.

\subsection{Domain\sphinxhyphen{}specific Textual Languages}
\label{\detokenize{background:domain-specific-textual-languages}}
\sphinxAtStartPar
A domain\sphinxhyphen{}specific language (DSL) for music is a textual programming language designed
expressly around making music with a computer (Wang 2008, 1).
.. citation (Wang, 2008, check page \sphinxhyphen{} replace with better DSL def citation if I find one FIX  (maybe swap to Lazzarini?)

\sphinxAtStartPar
Interestingly, the first example of programming computer music (that one might reasonably
consider as more than an audio experiment) used a music DSL, namely Max Matthew’s MUSIC I
language, created in 1957.
MUSIC I (originally refered to as simply MUSIC) was a domain specific language written in assembly
language for the IBM 704 mainframe at Bell Labs.
It was able to translate a high\sphinxhyphen{}level text language with musical abstractions to assembly code,
and could (through various intermediary steps) output digital audio.
MUSIC I was followed by various refinements by Matthews (Music II through V),
and by similar languages by others.
Its lineage continues to this day in the Csound language, still under active development and widely used,
and one with which I have extensive experience (Manning, 2013, 187\sphinxhyphen{}189).
.. citation (Manning, 2013)

\sphinxAtStartPar
While the source code of Csound piece, for example, is clearly a computer programs
(and would be recognizable as such to one familiar with programming)
the way in which it turns code into music would not likely be obvious at a glance to a programmer unfamiliar with music.
The language is, to a significant degree, designed around high\sphinxhyphen{}level abstractions suitable for particular ways
of creating a composition, and has a particular way in which it is run to make the final product.
Historically, running such a program meant rendering a piece to an audio file, but
with modern computers (and versions of Csound) the rendering can be done in realtime.
While originally these programs were not something with which a performer could interact while the music rendered,
facilties now exist in Csound for performers to interact with the programs while they play (Lazzarini, 2016).
.. citation (Lazzarini, 2013)

\sphinxAtStartPar
In addition to Csound, some other actively developed examples from this general family of language
include SuperCollider, ChucK, and Faust, each of which has a particular focus or approach to the problems
of computer music (Wang, 2008).
.. citation FIX

\sphinxAtStartPar
A notable advantage of a using a music DSL is that many of the hard
decisions that face the programmer have been made already.
The composer\sphinxhyphen{}programmer is not starting with a blank slate:
the language provides built\sphinxhyphen{}in abstractions ranging from
macro\sphinxhyphen{}structural concepts such as scores and sections to individual notes and beats.
Music DSLs thus significantly simplify the task of programming music and reduce
how much the composer\sphinxhyphen{}programmer must learn and program to begin making music (Lazzarini, 2017, 26).
.. citation (Lazzarini, 2017)
In Csound, for example, a program consists of an “orchestra” file, containing
programmatic instrument definitions, and a “score” file, containing a score
of musical events notated in Csound’s own data format.
These are used together to render a scored piece to audio,
either as an offline operation or as a realtime operation.

\sphinxAtStartPar
In the simple example in figure 1, we have a instrument playing a
short melody driven by the score.

\sphinxAtStartPar
With these built\sphinxhyphen{}in musical constructs, DSL’s are attractive to the composer\sphinxhyphen{}programmer,
but on the other hand, the tools\sphinxhyphen{}programmer is significantly more constrained than when
working in a general purpose programming language.
This can be frustrating for experienced programmers coming from general purpose languages,
who may wonder where their function calls and looping constructs went and how they can
express algorithms they already know in the abstractions provided by the language.
For example, in Csound one can program a form of recursion, but one of the techniques
for doing so involves creating instruments that play notes that in turn schedule notes (Lazzarini, 2016, 116).
.. citation (Lazzarini, 2016)
The use of the note as the fundamental
unit of computation (where a “note” is an instance of an instrument definition activated at
some time, for some duration) requires the tools\sphinxhyphen{}programmer to not only
understand the concept of recursion, but to also understand how to translate it
into an unusual syntax.

\sphinxAtStartPar
That said, music DSLs generally provide ways of \sphinxstyleemphasis{extending} the language with
a general purpose language, allowing the tools\sphinxhyphen{}programmer to add new abstractions to the DSL itself.
In Csound, for example, a tools\sphinxhyphen{}programmer may create a new opcode (essentially the equivalent
of a Csound class or function) using the C language,
compiling it such that it can be used in the same way as any built in opcode that comes with Csound
(Fitch, 2011, 581).
.. citation (Fitch, 2011)

\sphinxAtStartPar
It should also be noted that the ease with which composer\sphinxhyphen{}programmers can work
with DSLs has led to broad popularity in the music community, and this
in turn has led to many programmers creating publically available extensions, thus providing
a rich library of freely\sphinxhyphen{}available tools for the programmer to use.
Csound, for example, is still actively used and developed today, which is
remarkable for a language first developed in 1986, and now has thousands of objects available (Manning, 2013, 189).
.. citation (Manning, 2013)
If an extension is popular and useful enough, it may even find its way into the
main language or into official repositories of extensions.

\sphinxAtStartPar
So how does a music DSL such as Csound stack up with regard to our three high level requirements?
Certainly, we are given many high\sphinxhyphen{}level and convenient musically\sphinxhyphen{}meaningful abstractions.
Creating linear pieces according to a set score is straightforward.
Performer interaction is also possible in modern versions, though programming
an interactive system is somewhat cumbersome in that tasks that would require simple programming
in a general purpose language
must be done in an unusual manner to fit in the note\sphinxhyphen{}centered paradigm of Csound.
For example,  making a component to receive, parse, and translate MIDI input according
to some arbitrary rules requires making an “instrument” and having the
score turn on “always\sphinxhyphen{}on” notes (Lazzarini, 2106, 175).
.. citation (Lazzarini, 2016)
Clearly we are bending the built in abstractions to other purposes,
at the expense of easily comprehensible code.

>>>>

\usepackage{sphinxmessages}
\setcounter{tocdepth}{1}

@@

\sphinxAtStartPar
A side effect of the rapid progress in realtime audio computation is that many techniques
used principally in the context of \sphinxstyleemphasis{non\sphinxhyphen{}realtime} computer music have largely fallen out of the public light.
However, in the last decade computers have reached speeds such that some of the
tools and approaches previously associated only with non\sphinxhyphen{}realtime rendering have become
feasible even in a realtime scenario.
One of these, the one with which I am involved, is the programming of computer music in the
Lisp family of languages.
With its support for symbolic processing, recursion, and iteractive development,
Lisp has always been an elegant and productive way to represent and build music (CTN: Dannenberg 1997, 291).
However, the accepted wisdom has been that one does not try to do anything
where the same computer is generating significant amounts of digital audio in real time
while running a Lisp or similar high\sphinxhyphen{}level, garbage\sphinxhyphen{}collected language.
This is now not just possible, but practical even at latencies usable by serious instrumentalists
(i.e., sub 20 ms), allowing it be used on stage as well as in the studio.

\sphinxAtStartPar
I believe we are, as a result, at the cusp of a renaissance of the use of high\sphinxhyphen{}level programming languages in music.
JavaScript is used as an extension language in Max,
Python and Ruby are used for algorithmic composition,
and even Haskell is used in live\sphinxhyphen{}coding performances
(CITE:Lyon 2012, 13\sphinxhyphen{}16; Mclean and Dean 2018, 650\sphinxhyphen{}651, 597, 258).
And a new generation of computer
musicians is discovering the elegance, the productivity, and the joy of doing this work
with Lisp, a language that I and many others will argue is uniquely suited to expressing musical
abstractions in code.

\sphinxAtStartPar
The Scheme for Max project is, among other things, an attempt to make music programming in a Lisp
\sphinxhyphen{} specifically the Scheme dialect of Lisp \sphinxhyphen{} accessible and practical to the “regular” computer musician working in a “regular” studio.
Scheme for Max (S4M) provides an extension to the Max audio/visual programming platform
that embeds a Scheme interpreter, allowing the user to create fully functional Scheme programs that
interact with the rest of the Max environment, and which do so with high temporal accuracy.
S4M can be used to embed single\sphinxhyphen{}line programs directly in visual Max patches, or to build
programs of thousands of lines that do everything from sequencing MIDI
to controlling analog hardware and editing audio.
And it can be used for music of all sorts, from complex contemporary scored music, to interactive
live improvisation, to generative and stochastic algorithmic composition.

\sphinxAtStartPar
While I am far from the only one working on new ways to use Lisp in computer music (even in a realtime context),
I believe Scheme for Max achieves something unique.
By creating an enviroment thats runs seamlessly and flexibly inside the tremendously successful
Max platform (and thus also inside the even more tremendously successful Ableton Live digital audio workstation),
Scheme can be productively used in close tandem with both popular,
commercial music tools and the vast library of specialized, academic, and research tools available for Max.
And by the happy coincidence of Scheme’s syntax compability with Max message syntax, it can be even be used
in baby\sphinxhyphen{}steps \sphinxhyphen{} new programmers can add the power of Scheme to their Max patches with
only a few lines of code without even reaching for a text editor.
I believe Scheme’s consistent, minimal syntax and interactive development style provide an ideal
introduction to textual programming for the curious Max user.
With Scheme for Max, one is not forced to choose between the power and flexibility of Lisp and the
convenience and accessibility of commercial music software.

\sphinxAtStartPar
This thesis introduces the Scheme for Max project. I begin by discussing the landscape into which it fits
and its precendents and inspirations. I cover the motivations and goals for the project, discussing
my own musical and technical desires.
I examine its design and capabilities. And finally, I conclude by evaluating
how well it has succeeded, discussing its limitations, and outlining my future plans for the project.

\sphinxAtStartPar
For reasons of space, I have not covered every feature of the project, nor have I delved into the
low\sphinxhyphen{}level implementation details and the underlying C code. I have limited myself to discussion appropriate
to an interested computer musician, rather than a professional programmer or computer scientist.
That said, the project is open\sphinxhyphen{}source and freely available, so the
the interested reader is encouraged to consult the GitHub project page for further details if desired,
where they will find links to comprehensive documentation, help files, demo videos, and source code.
Code samples presented here in Scheme should be comprehensible to a new programmer,
but if the reader requires more help, I would point them to my “Learn Scheme For Max” online
e\sphinxhyphen{}book in which I introduce Scheme programming from first\sphinxhyphen{}principles, available again
from the main project page at (\sphinxurl{https://github.com/iainctduncan/scheme-for-max}).

@@

\sphinxAtStartPar
To begin, I would like to clarify exactly what I mean by “computer music programming”.
I am using the term to refer to programming in which both the tools used and the composition itself
are programmed on a computer.
I say this using the word “composition” in its broadest sense \sphinxhyphen{} the composition could
be a linearly scored piece of set length and content, or an interactive program with which
a performer interacts, but that contains prepared material of some kind.
This is distinct, for example, from commercial music tools (such as sequencers or digital
audio workstations) in which a computer program is run as the host environment, but the compositional
work itself is stored as static data within that program (for example in a binary format or
in MIDI files), and would not normally be considered a computer \sphinxstyleemphasis{program} itself.

\sphinxAtStartPar
Using this definition of computer music programming, we can draw a further distinction
between two kinds of programming and programmer:
we have programming that is specific to a composer or an artistic work;
and we have programming of tools and frameworks that will be potentially be used by many
people and in potentially many contexts.
For the purpose of this discussion, I will refer to the people doing these as the “composer\sphinxhyphen{}programmer”
\sphinxhyphen{} a term also used by Curtis Roads in “Composing Electronic Music” (CTN: Roads 2015, 341) \sphinxhyphen{}
and the “tools programmer” (my own terminology).
Of course, these frequently are the same person at different times.
This distinction is important, as the goals of these two programmers, and thus the
ideal design of supporting technology, can frequently be very different.
The composer\sphinxhyphen{}programmer likely wishes the programming tools to be as immediately convenient as possible to
an individual, while the tools\sphinxhyphen{}programmer may be
developing software for use by a larger community and may prefer design decisions
that favour long\sphinxhyphen{}term code reuse and team development at the expense of immediate convenience.
As we will see, the various computer music languages, platforms, and approaches
favour one of these hypothetical programmers to a greater or lesser degree.
Beyond these distinctions, I use the term “computer music programming” as broadly
as possible \sphinxhyphen{} I include as computer music programming platforms both graphic platforms with
which a beginner can be immediately productive as well as general purpose programming
languages that are the domain of experienced computer programmers.

\section{Requirements for a Computer Music Platform}
\label{\detokenize{background:requirements-for-a-computer-music-platform}}
\sphinxAtStartPar
Prior to examining the programming platforms, let us examine some
common requirements of the composer\sphinxhyphen{}programmer so that we can assess how the various
options meets these needs.
While the list of possible requirements is potentially long, I will group
them into three high level requirements that are personally important to me as
a composer and performer. Note that not all of these are necessarily fulfilled well, or at all,
by each of the possible platforms a composer might use \sphinxhyphen{} some offer
only partial support for a requirement, or perhaps even none at all (CTN: Lazzarini 2013, 97).

\subsection{Support for Musical Abstractions}
\label{\detokenize{background:support-for-musical-abstractions}}
\sphinxAtStartPar
To be maximally productive in producing musical works, the computer music language should provide
musically meaningful abstractions that support how the composer thinks about music.
Examples of these abstractions would be programmatic constructs for representing
notes, scores, instruments, voices, tempi, bars, beats, and the like.
The more of these that are supported in the language itself, the less the programmer
must decide and code themselves in order to make music.
Probably the most important of these abstractions are those concerning musical time
\sphinxhyphen{} a good computer music language must have an implementation of \sphinxstyleemphasis{logical time},
allowing the programmer to schedule events accurately in a way that makes sense
to the user but that also works for rendering or playing in real\sphinxhyphen{}time to a degree of accuracy
appropriate for music (CTN: Dannenberg 2018, 3).

\sphinxAtStartPar
However, it should be noted that the importance of these abstractions being readily provided
by a language varies widely amongst composer\sphinxhyphen{}programmers:
while some composers will be relieved not to have to decide how to implement
the concept of beats and tempi themselves, others may be expressly looking for platforms
that do not come with any assumptions of how one might think about musical abstractions.
Computer music programming languages thus also vary widely in their areas
of focus and the problems they attempt to solve for the programmer (CTN: Dannenberg 2018, 4).
For example, the Csound language comes with a built\sphinxhyphen{}in abstraction of a score system,
that in turn depends on abstractions representing beats and tempi (CTN: Lazzarini 2016, 157).
In contrast, the Max platform takes the design approach of giving the user
an almost blank canvas, deliberately avoiding any assumptions that the user
will make music of any particular style or in any particular way (CTN: Puckette 2002, 34).

@@

\sphinxAtStartPar
Again, the importance of this varies with the composer and the platform.
In fact, this was not even possible with early computer music languages such
as the MUSIC\sphinxhyphen{}N family, with which the process of rendering audio from a score file
took far longer than the duration of a piece (CTN: Wang 2017, 60\sphinxhyphen{}63).

\sphinxAtStartPar
However, since the advent of realtime\sphinxhyphen{}capable systems, this has become a
standard feature in computer music platforms.
In fact, in contrast to the Music\sphinxhyphen{}N languages, the Max platform was designed
to support performance\sphinxhyphen{}time interactivity first, with the ability to render audio only added later
as computers became capable of doing both (CTN: Puckette 2002, 33).
As someone who comes principally from a jazz and contemporary improvised music
background, being able to create complex systems that support realtime input is a high priority to me.

\subsection{Support for Complex Algorithms}
\label{\detokenize{background:support-for-complex-algorithms}}
\sphinxAtStartPar
Finally, the platform should provide support for programming algorithms of significant complexity.
It should be possible to make the platform, for example, do things that are impractical
or impossible for the human perfomer or for the composer working at a piano with score paper.
My personal goals include creating music that combines algorithmically generated
material with traditionally scored and improvised material.
For my algorithmic work, I want to explore techniques and forms that would be impractical
if one is individually entering notes, thus instead requiring programmatic implementation of
algorithms.
However, it is also important to me that a performer be able to interact with the algorithms.
That, to quote Miller Puckette on his original design goals for Max,
basic musical material should be “controllable physcially, sequentially, or algorithmically;
if they are to be controlled algorithmically, the inputs to the algorithm should
themselves be controllable in any way” (CTN: Puckette 1991, 68).

@@

\sphinxAtStartPar
I will briefly discuss each of these, listing various examples, but focusing on a representative tool from each family.
I will provide my observations and experiences of the advantages and disadvantages of each,
drawing both on the literature and on my personal experiences with tools from each category
over the last 25 years.

\subsection{Domain\sphinxhyphen{}Specific Textual Languages}
\label{\detokenize{background:domain-specific-textual-languages}}
\sphinxAtStartPar
A domain\sphinxhyphen{}specific language (DSL) for music is a textual programming language intended
expressly for making music with a computer (CTN: Wang 2017, 58).

\sphinxAtStartPar
The first historical example of programming computer music (that one might reasonably
consider as more than an audio experiment) used a music DSL, namely Max Matthew’s MUSIC I
language, created in 1957.
MUSIC I (originally refered to as simply MUSIC) was a domain specific language written in assembly
language for the IBM 704 mainframe at Bell Labs.
It was able to translate a high\sphinxhyphen{}level textual language with musical abstractions to assembly code,
and could (through various intermediary steps) output digital audio.
MUSIC I was followed by various refinements by Matthews (Music II through V),
and by similar languages by others.
Its lineage continues to this day in the Csound language, still under active development and widely used,
and one with which I have extensive experience. (CTN: Manning 2013, 187\sphinxhyphen{}189).

\sphinxAtStartPar
While the source code of Csound piece, for example, is clearly a computer programs
(and would be recognizable as such to one familiar with programming)
the way in which it turns code into music would not likely be obvious at a glance to a programmer unfamiliar with music.
The language is, to a significant degree, designed around high\sphinxhyphen{}level abstractions suitable for particular ways
of creating a composition, and has a particular way in which it is run to make the final product.
Historically, running such a program meant rendering a piece to an audio file, but
with modern computers (and versions of Csound) the rendering can be done in realtime.
While originally these programs were not something with which a performer could interact while the music rendered,
facilties now exist in Csound for performers to interact with the programs while they play (CTN: Lazzarini 2016, 171\sphinxhyphen{}179).

\sphinxAtStartPar
In addition to Csound, some other actively developed examples from this general family of language
include SuperCollider, ChucK, and Faust, each of which has a particular focus or approach to the problems
of computer music (CTN: Wang 2017, 69\sphinxhyphen{}72; Lazzarini 2017, 41\sphinxhyphen{}42).

\sphinxAtStartPar
A notable advantage of a using a music DSL is that many of the hard
decisions that face the programmer have been made already.
The composer\sphinxhyphen{}programmer is not starting with a blank slate:
the language provides built\sphinxhyphen{}in abstractions ranging from
macro\sphinxhyphen{}structural concepts such as scores and sections to individual notes and beats.
Music DSLs thus significantly simplify the task of programming music and reduce
how much the composer\sphinxhyphen{}programmer must learn and program to begin making music (CTN: Lazzarini 2017, 26).
In Csound, for example, a program consists of an “orchestra” file, containing
programmatic instrument definitions, and a “score” file, containing a score
of musical events notated in Csound’s own data format. (These files
may be merged into one containing “csd” file, but the distinction still holds.)
These are used together to render a scored piece to audio,
either as an offline operation or as a realtime operation.
A sample of Csound code is shown below, with an instrument playing a
short scale driven by the score.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{o}{\PYGZlt{}}\PYG{n}{CsoundSynthesizer}\PYG{o}{\PYGZgt{}}
\PYG{o}{\PYGZlt{}}\PYG{n}{CsInstruments}\PYG{o}{\PYGZgt{}}

\PYG{k+kd}{instr}\PYG{+w}{ }\PYG{n+nf}{1}
\PYG{+w}{  }\PYG{c+c1}{; take pitch as midi note from param 4}
\PYG{+w}{  }\PYG{k+kt}{k}\PYG{n}{frq}\PYG{+w}{  }\PYG{n+nb}{mtof}\PYG{+w}{  }\PYG{n+nv+vi}{p4}
\PYG{+w}{  }\PYG{c+c1}{; saw wave}
\PYG{+w}{  }\PYG{k+kt}{a}\PYG{n}{sig}\PYG{+w}{  }\PYG{n+nb}{vco2}\PYG{+w}{  }\PYG{l+m+mf}{0.8}\PYG{p}{,}\PYG{+w}{ }\PYG{k+kt}{k}\PYG{n}{frq}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mi}{0}
\PYG{+w}{  }\PYG{c+c1}{; an ADSR envelope}
\PYG{+w}{  }\PYG{k+kt}{a}\PYG{n}{env}\PYG{+w}{  }\PYG{l+m+mf}{0.01}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mf}{0.3}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mf}{0.7}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mf}{0.2}
\PYG{+w}{  }\PYG{c+c1}{; apply env and output in stereo}
\PYG{+w}{  }\PYG{n+nb}{outs}\PYG{+w}{  }\PYG{k+kt}{a}\PYG{n}{sig}\PYG{+w}{ }\PYG{o}{*}\PYG{+w}{ }\PYG{k+kt}{a}\PYG{n}{env}\PYG{p}{,}\PYG{+w}{ }\PYG{k+kt}{a}\PYG{n}{sig}\PYG{+w}{ }\PYG{o}{*}\PYG{+w}{ }\PYG{k+kt}{a}\PYG{n}{env}
\PYG{k+kd}{endin}

\PYG{o}{\PYGZlt{}}\PYG{o}{/}\PYG{n}{CsInstruments}\PYG{o}{\PYGZgt{}}
\PYG{o}{\PYGZlt{}}\PYG{n}{CsScore}\PYG{o}{\PYGZgt{}}

\PYG{c+c1}{;  time   dur  midi\PYGZhy{}note\PYGZhy{}num}
\PYG{k+kt}{i}\PYG{n}{1}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{+w}{      }\PYG{l+m+mi}{1}\PYG{+w}{    }\PYG{l+m+mi}{60}
\PYG{k+kt}{i}\PYG{n}{1}\PYG{+w}{ }\PYG{l+m+mi}{1}\PYG{+w}{      }\PYG{l+m+mi}{1}\PYG{+w}{    }\PYG{l+m+mi}{62}
\PYG{k+kt}{i}\PYG{n}{1}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{+w}{      }\PYG{l+m+mi}{1}\PYG{+w}{    }\PYG{l+m+mi}{64}
\PYG{k+kt}{i}\PYG{n}{1}\PYG{+w}{ }\PYG{l+m+mi}{3}\PYG{+w}{      }\PYG{l+m+mi}{1}\PYG{+w}{    }\PYG{l+m+mi}{65}
\PYG{k+kt}{i}\PYG{n}{1}\PYG{+w}{ }\PYG{l+m+mi}{4}\PYG{+w}{      }\PYG{l+m+mi}{1}\PYG{+w}{    }\PYG{l+m+mi}{67}

\PYG{o}{\PYGZlt{}}\PYG{o}{/}\PYG{n}{CsScore}\PYG{o}{\PYGZgt{}}
\PYG{o}{\PYGZlt{}}\PYG{o}{/}\PYG{n}{CsoundSynthesizer}\PYG{o}{\PYGZgt{}}
\end{sphinxVerbatim}

\sphinxAtStartPar
With their built\sphinxhyphen{}in musical abstractions, DSL’s are attractive to the composer\sphinxhyphen{}programmer,
but on the other hand, the tools\sphinxhyphen{}programmer is significantly more constrained than when
working in a general purpose programming language.
This can be frustrating for experienced programmers coming from general purpose languages,
who may wonder where their function calls and looping constructs went and how they can
express the algorithms with which they are familiar in the unusual abstractions provided by the language.
For example, in Csound one can program a form of recursion, but one of the techniques
for doing so involves creating instruments that play notes that in turn schedule notes (CTN: Lazzarini 2016, 116).
The use of the note as the fundamental
unit of computation (where a “note” is an instance of an instrument definition activated at
some time, for some duration) requires the tools\sphinxhyphen{}programmer to not only
understand the concept of recursion, but to also understand how to translate it
into this unusual syntax.

\sphinxAtStartPar
That said, music DSLs generally provide ways of \sphinxstyleemphasis{extending} the language with
a general purpose language, allowing the tools\sphinxhyphen{}programmer to add new abstractions to the DSL itself.
In Csound, for example, a tools\sphinxhyphen{}programmer may create a new \sphinxstyleemphasis{opcode} (essentially the equivalent
of a Csound class or function) using the C language,
compiling it such that it can be used in the same way as any built in opcode that comes with Csound
(CTN: ffitch 2011b, 581).

\sphinxAtStartPar
It should also be noted that the ease with which composer\sphinxhyphen{}programmers can work
with DSLs has led to broad popularity in the music community, and this
in turn has led to many programmers creating publically available extensions, thus providing
a rich library of freely\sphinxhyphen{}available tools for the programmer to use.
Csound, for example, is still actively used and developed today, which is
remarkable for a language first developed in 1986, and now has thousands of objects available (CTN Manning 2013, 189).
If an extension is popular and useful enough, it may even find its way into the
main language or into official repositories of extensions.
